---
title: "Plant functional type (PFT) comparison models (inputs for Fig. 3)"
output: html_document
---

This notebook fits and evaluates the two alternative models in which plant
functional type (IGBP class) replaces the trait / functional-diversity terms:

1. `GPP ~ LAI + Temperature + Precipitation + aridity_index + IGBP + (1 | ecoprovince)`
2. `GPP ~ LAI + Temperature + Precipitation + aridity_index * IGBP + (1 | ecoprovince)`

Both use the same Gaussian spatial correlation structure
(`corGaus(~ latitude + longitude, nugget = TRUE)`) as the other models in this
folder, and are compared with the baseline model (partial R2 and delta AIC).

* Inputs: `../data/analysis_ready_tables/data.csv`,
  `LME_GPP_LAI_T_P_AI_ecoprovince_SA.rds` (baseline, from `01_fit_composition_models.R`)
* Outputs: `LME_GPP_LAI_T_P_AI_IGBP_ecoprovince_SA.rds`,
  `LME_GPP_LAI_T_P_AI_IGBP_AI_ecoprovince_SA.rds`,
  `results/LME_GPP_LAI_T_P_AI_IGBP_ecoprovince_SA_results.csv`,
  `results/LME_GPP_LAI_T_P_AI_IGBP_AI_ecoprovince_SA_results.csv`
* Run order: after `09_evaluate_models.Rmd`, before `11_plot_fig3.py`
  (knit with the working directory set to this folder).
* Feeds: Fig. 3, "+ PFT" panel (`11_plot_fig3.py`).

```{r setup, include=FALSE}
library(nlme)
library(performance)
library(knitr)
library(dplyr)

dir.create("results", showWarnings = FALSE)
```

# Fit model: GPP ~ LAI + Temperature + Precipitation + aridity_index + IGBP + (1 | ecoprovince)

Baseline model (LAI + T + P + AI) plus PFT (IGBP as a categorical fixed effect).
The model is fitted only if the saved `.rds` does not already exist.

```{r fit-model}
# Load the analysis-ready table
data <- read.csv("../data/analysis_ready_tables/data.csv")

# IGBP as a factor
data$IGBP <- as.factor(data$IGBP)

model_file <- "LME_GPP_LAI_T_P_AI_IGBP_ecoprovince_SA.rds"

if (file.exists(model_file)) {
  model_IGBP <- readRDS(model_file)
} else {
  formula <- as.formula("GPP ~ LAI + Temperature + Precipitation + aridity_index + IGBP")
  cor_struct <- corGaus(form = ~ latitude + longitude, nugget = TRUE)
  model_IGBP <- lme(fixed = formula,
                    random = ~ 1 | ecoprovince,
                    data = data,
                    method = "ML",
                    correlation = cor_struct)
  saveRDS(model_IGBP, model_file)
}

# Baseline model (without IGBP), the reference for partial R2 and delta AIC
model_LAI <- readRDS("LME_GPP_LAI_T_P_AI_ecoprovince_SA.rds")
```

# Model performance and comparison with the baseline (LAI + T + P + AI)

```{r performance}
perf_IGBP <- model_performance(model_IGBP, estimator = "ML")
perf_LAI  <- model_performance(model_LAI,  estimator = "ML")

R2_marginal_LAI    <- perf_LAI$R2_marginal
R2_marginal_IGBP   <- perf_IGBP$R2_marginal
R2_conditional_IGBP <- ifelse(is.na(perf_IGBP$R2_conditional),
                              R2_marginal_IGBP, perf_IGBP$R2_conditional)
ICC_IGBP  <- ifelse(is.null(perf_IGBP$ICC), 0, perf_IGBP$ICC)

# Overall significance of IGBP (likelihood-ratio test against the baseline)
anova_res <- anova(model_LAI, model_IGBP)
p_LRT_IGBP <- anova_res[2, "p-value"]

results_df <- data.frame(
  model                = "LAI_T_P_AI_IGBP",
  n_IGBP_levels        = nlevels(data$IGBP),
  R2_marginal          = R2_marginal_IGBP,
  R2_conditional       = R2_conditional_IGBP,
  partial_R2_LAI       = R2_marginal_LAI,
  partial_R2_IGBP      = R2_marginal_IGBP - R2_marginal_LAI,
  ICC                  = ICC_IGBP,
  AIC                  = perf_IGBP$AIC,
  delta_AIC_vs_null    = perf_IGBP$AIC - perf_LAI$AIC,
  RMSE                 = perf_IGBP$RMSE,
  LRT_p_value_vs_null  = p_LRT_IGBP
)

write.csv(results_df,
          "results/LME_GPP_LAI_T_P_AI_IGBP_ecoprovince_SA_results.csv",
          row.names = FALSE)
kable(results_df, format = "markdown", digits = 4)
```

# Fit model: GPP ~ LAI + T + P + AI + IGBP + IGBP:AI + (1 | ecoprovince)

IGBP (categorical) interacting with aridity_index, i.e. a separate aridity
slope for each IGBP class. The model keeps the IGBP main effect (class-specific
intercepts) and the IGBP:aridity_index term (class-specific slopes).

```{r fit-igbp-ai}
model_file_int <- "LME_GPP_LAI_T_P_AI_IGBP_AI_ecoprovince_SA.rds"

if (file.exists(model_file_int)) {
  model_IGBP_AI <- readRDS(model_file_int)
} else {
  formula_int <- as.formula(
    "GPP ~ LAI + Temperature + Precipitation + aridity_index * IGBP"
  )
  cor_struct <- corGaus(form = ~ latitude + longitude, nugget = TRUE)
  model_IGBP_AI <- lme(fixed = formula_int,
                       random = ~ 1 | ecoprovince,
                       data = data,
                       method = "ML",
                       correlation = cor_struct)
  saveRDS(model_IGBP_AI, model_file_int)
}
```

## Performance comparison (Null vs +IGBP vs +IGBP*AI)

```{r perf-int}
perf_int <- model_performance(model_IGBP_AI, estimator = "ML")

compare_df <- data.frame(
  model          = c("Null", "Null + IGBP", "Null + IGBP*AI"),
  R2_marginal    = c(perf_LAI$R2_marginal,  perf_IGBP$R2_marginal,  perf_int$R2_marginal),
  R2_conditional = c(perf_LAI$R2_conditional, perf_IGBP$R2_conditional, perf_int$R2_conditional),
  AIC            = c(perf_LAI$AIC, perf_IGBP$AIC, perf_int$AIC),
  RMSE           = c(perf_LAI$RMSE, perf_IGBP$RMSE, perf_int$RMSE)
) %>%
  mutate(delta_AIC_vs_null = AIC - perf_LAI$AIC)

write.csv(compare_df,
          "results/LME_GPP_LAI_T_P_AI_IGBP_AI_ecoprovince_SA_results.csv",
          row.names = FALSE)
kable(compare_df, format = "markdown", digits = 4)
```
